---
title: "NectarExperiment_RMD"
output: html_document
date: "2023-09-11"
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
#setwd("NectarExperiment_Analysis/")
```
## Monarch Nectar Experiment Analysis
### Introduction:
####Monarchs feed on different flowers in the wild. I sought to determine how these flowers they feed on affect the fitness of the butterfly during the fall migratory period, looking at lean mass (proxy of protein), fat mass, and water mass.
####I selected 6 difference species to test in a greenhouse and raised butterflies on these flowers for 7 days (6 butterflies in the beginning were taken to 10 days but the trials were shortened). A few trends have been noted in the literature: Monarchs appear to accumulate fat for their fall migration, but we are unsure where this starts and ends (i.e. do they accumulate fat in Canada, or later when they get to the southern US?). There are currently 3 different hypotheses (based in differences in observations) about the timing of fat accumulation:
#### 1. Beall 1948: Monarch butterflies accumulate fat before they begin migrating, and during migration they feed to maintain or top up fat levels (so high fat levels early in migration, mid-high as they travel south).
####2.Brower et al 1985: Monarch butterflies do not accumulate any appreciable amount of fat until they reach the southern United States and Northern Mexico, where they begin to accumulate fat and store for the wintering period.
####3. Gibo & McCurdy 1993: Based on the differences between Beall & Brower --> Early in the migration period, monarch butterflies have low fat levels as they have not had time to feed on nectar and accumulate fat. As the migration period progresses, they have had more time to feed and have higher fat levels. However, butterflies that are late in the migratory period will have low fat levels as the quality of nectar plants drops and weather becomes less favourable, causing them to deplete reserves to make the migration (so early emergers have low-mid fat, but gain fat in the mid-migratory period, and late migrants (which have presumably emerged late) will have low fat).
#### My experiment: I raised freshly emerged butterflies on different nectar plants for 7 days each. This eliminates the "amount of time to feed" portion of Gibo and McCurdy's hypothesis (all butterflies will have 7 days)




### Diagnostics of models

```{r,echo=FALSE}
### Statistical analysis script
### Written by M. Veselovsky
### Last modified 2023-09-11

#Load data and libraries
summarydat<-read.csv("processed/summarydat.csv")
data = summarydat
library(lmerTest)
library(car)
library(effects)
library(ggplot2)
library(emmeans)
library(multcomp)
library(sjPlot)
library(sjmisc)
library(ggplot2)
library(ggeffects)
library(ggbeeswarm)
library(performance) #for check_model() assumptions function


# create separate working databases for greenhouse plants, goldenrod (for GH-Field comparison), and field plants
GH_data = subset(summarydat, summarydat$ExpLoc == "GH") # data from greenhouse
GH_F_data = summarydat[which(summarydat$Plant=="SolAlt"),] #all data on goldenrod, for GH-Field comparison
Field_data = subset(summarydat,summarydat$ExpLoc == "Field") #data from field
SolAlt_F_data = subset(Field_data,Field_data$Plant == "SolAlt") #Subset SolAlt in field to check for enclosure effects
```

```{r,echo=FALSE}


########## 1. GREEHOUSE DATA ANALYSIS ##############
################## 1.1 Normality check ###############
opar <- par(mfrow = c(1, 1))

#Check for normality of the butterflies raw weights on day 7 of their respective trials
weight_distr <- ggplot(GH_data, aes(x = RawWeight_day7))
weight_distr <- weight_distr +
  # add data density smooth
  geom_density() +
  # add rug (bars at the bottom of the plot)
  geom_rug() +
  # add black semitransparent histogram
  geom_histogram(aes(y = ..density..),
                 color = "black",
                 alpha = 0.3) +
  # add normal curve in red, with mean and sd from fklength
  stat_function(fun = dnorm,
                args = list(
                  mean = mean(GH_data$RawWeight_day7),
                  sd = sd(GH_data$RawWeight_day7)
                ),
                color = "red")

#display graph
weight_distr
qqnorm(GH_data$RawWeight_day7)
qqline(GH_data$RawWeight_day7)


#Check for normality of fat weight in butterflies after trial finished
fat_distr <- ggplot(GH_data, aes(x = DryFatMass))
fat_distr <- fat_distr +
  # add data density smooth
  geom_density() +
  # add rug (bars at the bottom of the plot)
  geom_rug() +
  # add black semitransparent histogram
  geom_histogram(aes(y = ..density..),
                 color = "black",
                 alpha = 0.3) +
  # add normal curve in red, with mean and sd from fklength
  stat_function(fun = dnorm,
                args = list(
                  mean = mean(GH_data$DryFatMass),
                  sd = sd(GH_data$DryFatMass)
                ),
                color = "red")

#display graph
fat_distr
qqnorm(GH_data$DryFatMass)
qqline(GH_data$DryFatMass)


# Formal test for normality (Shapiro)
shapiro.test(GH_data$RawWeight_day7) #weights normal
shapiro.test(GH_data$DryFatMass) #fat non-normal

# boxplots of raw weight on day 7 by plant species in the greenhouse
myplot<-ggplot(data=GH_data, aes(x=Plant, y=RawWeight_day7),na.action=na.exclude)
myplot+geom_boxplot(notch=FALSE)


```
###Exploratory plots
```{r,echo=FALSE}
#########plots for sex differences
myplot<-ggplot(data=GH_data, aes(x=Sex, y=DryFatMass),na.action=na.exclude)
myplot+geom_boxplot(notch=FALSE)


myplot<-ggplot(data=GH_data, aes(x=Sex, y=WaterMass),na.action=na.exclude)
myplot+geom_boxplot(notch=FALSE)

myplot<-ggplot(data=GH_data, aes(x=Sex, y=DryLeanMass),na.action=na.exclude)
myplot+geom_boxplot(notch=FALSE)
```


```{r,echo=FALSE}
#################### 1.2 MODEL BUILDING - LMER ###########################

# Check correlation of starting weight and forewing length (to see if I need both in the model)
# The a-priori expectation is that they will be highly correlated, in which case I think forewing length is
# the ideal variable to retain as it does not change with adult age or time of day (feeding could change weight slightly).
# Weights could fluctuate depending on the time they were taken
# since emergence (metabolism as they have not eaten, and drying time for the fluid they expel
# as they eclose from their chrysalises)

fwl_day0w_lm = lm(RawWeight_day0 ~ ForewingLength, data=GH_data)
Anova(fwl_day0w_lm, type=3)# Significantly related, so I will keep forewing length and not use weight
plot(allEffects(fwl_day0w_lm))
summary(fwl_day0w_lm)

```

#### Model of the wet weight of the butterfly as a function the sex of the individual, forewing length, emergence date, and plant species + total surface area of flowers
```{r}

# Estimate the change in raw weight on day 7 of trial based on predictor variables
weight_lm = lm(RawWeight_day7~ Plant + Sex + ForewingLength + EmergDate + TotalSA, data=GH_data)
```


```{r,echo=FALSE}
#Test for collinearity among predictor variables (GVIF<4 for continuous acceptable,
# GVIF^(1/(2*df)) < 2 acceptable for categorical)
vif(weight_lm) #all good
opar <- par(mfrow = c(2, 2))

plot(weight_lm)
plot(weight_lm, resid(., scaled=TRUE) ~ fitted(.), abline = 0,pch=16,xlab="Fitted values",ylab="Standardised residuals")
Anova(weight_lm, type=3)
summary(weight_lm)

effects_info = allEffects(weight_lm)
effects_info
plot(effects_info)


############ 1.3 Multiple comparisons test

emmeans(weight_lm, list(pairwise~Plant), adjust="tukey")

######## 2. COMPARISON OF GH TO FIELD DATA (USING GOLDENROD) ##############
#Create a database of field and gh goldenrod to compare experiments
GH_F_data = summarydat[which(summarydat$Plant=="SolAlt"),]
summary(GH_F_data)

#Check for normality on raw day 7 weights
hist(GH_F_data$RawWeight_day7)


### linear model to compare GH method to field method (ExpLoc)
gh_f_lm = lm(RawWeight_day7~ Sex + ForewingLength + ExpLoc, data=GH_F_data)
Anova(gh_f_lm, type=3)

plot(allEffects(gh_f_lm))
summary(gh_f_lm)


################ 3. PLANT SURFACE AREA DIFFERENCES BETWEEN SPECIES ############

########### 3.1 SA - GREENHOUSE COMPARISON
## Histogram of surface area by plant species
myplot<-ggplot(data=GH_data, aes(x=Plant, y=log(TotalSA)),na.action=na.exclude)
myplot+geom_boxplot(notch=TRUE)
stripchart(TotalSA ~ Plant, data=GH_data,              # Data
                      method = "jitter", # Random noise
                      pch = 19,          # Pch symbols
                      col = 4,           # Color of the symbol
                      add = TRUE) 

#### Compare surface area of the greenhouse plants
## scatter plot by plant species
ggplot(data = GH_data) +
  aes(y = Plant, x = log(TotalSA)) +
  geom_beeswarm() +
  coord_flip()
boxplot(TotalSA~Plant,data=GH_data)
#Model 
SA_comparison_GH = lm(TotalSA ~ Plant, data=GH_data)
summary(SA_comparison_GH)

#model effects
plot(allEffects(SA_comparison_GH), ylab="Floral Surface Area")
Anova(SA_comparison_GH)


########### 3.2 SA - FIELD SPECIES
ggplot(data = Field_data) +
  aes(y = Plant, x = TotalSA) +
  geom_beeswarm() +
  coord_flip()

ggplot(data = Field_data) +
  aes(y = Plant, x = log(TotalSA)) +
  geom_beeswarm() +
  coord_flip()

myplot<-ggplot(data=Field_data, aes(x=Plant, y=TotalSA),na.action=na.exclude)
myplot+geom_boxplot(notch=TRUE)

SA_field_lm = lm(TotalSA~Plant,data=Field_data)
Anova(SA_field_lm)
plot(allEffects(SA_field_lm))


###################### 4 FIELD PLANTS & COMPARISONS #####################

# Because greenhouse SolAlt (Tall goldenrod) had a different effect from field SolAlt,
# I will specifically compare the two field species (EutMac/Joe-pye weed & SolAlt)
# as they cannot be pooled with the greenhouse data
summary(Field_data)


##################### 4.1 CHECK FOR ENCLOSURE EFFECTS ####################
# Check for enclosure effects with just SolAlt data (SolAlt had two different types
# of enclosures in the trial)
encl_lm = lm(RawWeight_day7~ Sex + ForewingLength + EmergDate + EnclCol, data=SolAlt_F_data)

Anova(encl_lm) #EnclCol least significant, remove and compare
encl_reduced = lm(RawWeight_day7~ Sex + ForewingLength + EmergDate, data=SolAlt_F_data)
anova(encl_reduced, encl_lm) # more complex (with EnclCol) is not significantly better, leave out EnclCol

Anova(encl_reduced) #EmergDate least significant, remove and compare
encl_reduced2 = lm(RawWeight_day7~ Sex + ForewingLength, data=SolAlt_F_data)
anova(encl_reduced2, encl_reduced) # more complex (with EmergDate) is not significantly better
Anova(encl_reduced2)

#Sex least significant, remove and compare
encl_reduced3 = lm(RawWeight_day7~ ForewingLength, data=SolAlt_F_data)
anova(encl_reduced3, encl_reduced2) # more complex (with EmergDate) is not significantly better


#################### 4.2 COMPARE FIELD SPECIES ########################
# I had a limited number of field plots (only 3 plots of JPW)
# so the first set of field trials had 2 females per enclosure for EutMac,
# and a mix of double females, single males, and single females for SolAlt.
# For the second cohort, I put additional double females and males for both species
# and added a single female and single male for EutMac. Unfortunately, I did not get
# more male replicates for EutMac so I will exclude males from the field analysis
females_data = subset(Field_data,Field_data$Sex == "F")

#### build model using females, without EnclCol
females_data$NumButterflies = as.factor(females_data$NumButterflies)
field_lm = lm(RawWeight_day7~ Plant + ForewingLength + NumButterflies + TotalSA, data=females_data)
Anova(field_lm, type=3)
summary(field_lm)
plot(allEffects(field_lm))
summary(females_data)

field_fat = lm(RawWeight_day7~ Plant + ForewingLength + NumButterflies + TotalSA, data=females_data)

## remove plant, least significant term, and compare
#field_reduced = lm(RawWeight_day7~ ForewingLength + NumButterflies, data=females_data)

# anova(field_reduced, field_lm) #more complex is not significantly better
# Anova(field_reduced) #remove numButterflies
# 
# field_reduced2 = lm(RawWeight_day7~ ForewingLength, data=females_data)
# anova(field_reduced2,field_reduced)
# Anova(field_reduced2)



############### 5. FAT EXTRACTION ANALYSES ##########################



############### 5.1 Exploration ##################################
opar <- par(mfrow = c(1, 1))
boxplot(WaterMass~Plant,data=data,ylab="Body water content (g)")
water_lm = lm(WaterMass~Plant,data=data)
summary(water_lm)

ggplot(data, aes(x = Plant, y = WaterMass, color = ExpLoc)) +  # ggplot function
  geom_point()

scatterplot(DryFatMass~RawWeight_day7,data=data, ylab="Body fat (g)",xlab="Body mass - wet (g)")
fat_lm =lm(DryFatMass~RawWeight_day7,data=data)
summary(fat_lm)

ggplot(data, aes(x = RawWeight_day7, y = log(DryFatMass), color = ExpLoc)) +  # ggplot function
  geom_point()+
  geom_smooth(method=lm , color="red", se=TRUE) +

scatterplot(RelDryFat~RawWeight_day7,data=data,ylab="% body fat",xlab="Body mass - dry (g)")
scatterplot(RelDryFat~DryMass,data=data,ylab="% body fat",xlab="Body mass - dry (g)")

ggplot(data, aes(x = Plant, y = RelDryFat, color = ExpLoc)) +  # ggplot function
  geom_point()


relfat_lm = lm(RelDryFat~RawWeight_day7,data=data)
summary(relfat_lm)

scatterplot(DryMass~RawWeight_day7,data=data,ylab="Dry body mass (g)",xlab="Body mass - wet (g)")
drymass_lm = lm(DryMass~RawWeight_day7,data=data)
summary(drymass_lm)

scatterplot(DryFatMass~DryLeanMass,data=data,ylab="Body fat (g)",xlab="Lean mass (g)")

scatterplot(DryMass~EmergDate,data=data)


########################## 5.2 GREENHOUSE FAT MODELS ####################################
gh_fat_lm = lm(DryFatMass~ Plant + Sex + ForewingLength + EmergDate + TotalSA, data=GH_data)
gh_fat_lm_log = lm(log(DryFatMass)~ Plant + Sex + ForewingLength + EmergDate + TotalSA, data=GH_data)


#Test for collinearity among predictor variables (GVIF<4 for continuous acceptable,
# GVIF^(1/(2*df)) < 2 acceptable for categorical)
vif(gh_fat_lm)

opar <- par(mfrow = c(2, 2))
plot(gh_fat_lm)
plot(gh_fat_lm_log)

check_model(gh_fat_lm)
check_model(gh_fat_lm_log)

# Anova(gh_fat_lm)
# Anova(gh_fat_lm_log)
# 
# summary(gh_fat_lm_log)
# 
# pred_gh_fat_lm=allEffects(gh_fat_lm)
# pred_gh_fat_log=allEffects(gh_fat_lm_log)


# plot(pred_gh_fat_lm)
# plot(pred_gh_fat_log)
# Anova(gh_fat_lm_log)

predict_plant = predictorEffect("Plant",gh_fat_lm_log)

str(predict_plant)
predict_plant$fit = exp(predict_plant$fit)
predict_plant$upper=exp(predict_plant$upper)
predict_plant$lower=exp(predict_plant$lower)
plot(predict_plant,ylab ="Fat mass (g)")
plot(allEffects(gh_fat_lm_log))

emmeans(gh_fat_lm_log, list(pairwise~Plant), adjust="tukey")


################### GLM Using Gamma distribution for Fat #######################
gh_fat_lm_gm = glm(DryFatMass ~ Sex + Plant + ForewingLength + EmergDate + TotalSA, data=GH_data,family="Gamma")
opar <- par(mfrow = c(2, 2))
plot(gh_fat_lm_gm)

Anova(gh_fat_lm_gm)
summary(gh_fat_lm_gm)
pred_gh_fat_gm=allEffects(gh_fat_lm_gm)
plot(pred_gh_fat_gm)

str(pred_gh_fat_gm$Plant)
str(pred_gh_fat_gm$Plant$response)
str(pred_gh_fat_gm$fit)
str(pred_gh_fat_gm)

summary(gh_fat_lm_gm)
str(pred_gh_fat_gm)
plot(pred_gh_fat_gm)
check_model(gh_fat_lm_gm)

plot(pred_gh_fat_gm)
emmeans(gh_fat_lm_gm, list(pairwise~Plant), adjust="tukey")

## Plot of effects using ggplot glm
ggplot(GH_data, aes(EmergDate, DryFatMass)) + 
  geom_smooth(method = "glm", formula = DryFatMass~ Plant + Sex + ForewingLength + EmergDate + TotalSA, colour = "black",
              linetype = 2, fill = "gray80", alpha = 0.2,
              method.args = list(family = gamma)) +
  geom_rug(sides = "b") +
  theme_bw() +
  labs(y = "Fat weight (g)", x = "Plant") +
  theme(text = element_text(size = 16),
        plot.margin = margin(50, 50, 50, 50),
        axis.title.x = element_text(vjust = -8),
        axis.title.y = element_text(vjust = 10),
        plot.title = element_text(vjust = 8))






########## relative fat content (%)
gh_relfat_lm = lm(RelDryFat~ Plant + EmergDate + TotalSA, data=GH_data)
vif(gh_relfat_lm)
Anova(gh_relfat_lm)

emmeans(gh_relfat_lm, list(pairwise~Plant), adjust="tukey")



######################## 5.3 FIELD FAT ANALYSIS #######################
opar <- par(mfrow = c(2, 2))
field_fat_lm = lm(DryFatMass~ EmergDate + TotalSA, data=Field_data)
Anova(field_fat_lm)
plot(allEffects(field_fat_lm))
vif(field_fat_lm)

opar <- par(mfrow = c(2, 2))
plot(field_fat_lm)

field_fat_glm = glm(DryFatMass~ EmergDate, data=Field_data,family="Gamma")
Anova(field_fat_glm)

plot(allEffects(field_fat_glm))



#################### 6. LEAN MASS (PROTEIN) ANALYSIS ###############

lean_distr <- ggplot(GH_data, aes(x = DryLeanMass))
lean_distr <- lean_distr +
  # add data density smooth
  geom_density() +
  # add rug (bars at the bottom of the plot)
  geom_rug() +
  # add black semitransparent histogram
  geom_histogram(aes(y = ..density..),
                 color = "black",
                 alpha = 0.3) +
  # add normal curve in red, with mean and sd from fklength
  stat_function(fun = dnorm,
                args = list(
                  mean = mean(GH_data$DryLeanMass),
                  sd = sd(GH_data$DryLeanMass)
                ),
                color = "red")

#display graph
lean_distr
#generate the QQ plot
qqnorm(GH_data$DryLeanMass)
qqline(GH_data$DryLeanMass)

# 
# ggplot(data, aes(x = RawWeight_day7, y = DryLeanMass, color = ExpLoc)) +  # ggplot function
#   geom_point()+
#   geom_smooth(method=lm , color="red", se=TRUE)

gh_prot_lm = lm(DryLeanMass ~ Sex + Plant + ForewingLength + EmergDate, data=GH_data)
opar <- par(mfrow = c(2, 2))
plot(gh_prot_lm)

check_model(gh_prot_lm)
Anova(gh_prot_lm)

pred_prot_lm = allEffects(gh_prot_lm)
plot(pred_prot_lm)
emmeans(gh_prot_lm, list(pairwise~Plant), adjust="tukey")




##################### WATER ############################

water_distr <- ggplot(GH_data, aes(x = WaterMass))
water_distr <- water_distr +
  # add data density smooth
  geom_density() +
  # add rug (bars at the bottom of the plot)
  geom_rug() +
  # add black semitransparent histogram
  geom_histogram(aes(y = ..density..),
                 color = "black",
                 alpha = 0.3) +
  # add normal curve in red, with mean and sd from fklength
  stat_function(fun = dnorm,
                args = list(
                  mean = mean(GH_data$WaterMass),
                  sd = sd(GH_data$WaterMass)
                ),
                color = "red")

#display graph
water_distr
qqnorm(GH_data$DryLeanMass)
qqline(GH_data$DryLeanMass)

gh_water_lm = lm(WaterMass ~ Sex + Plant + ForewingLength + EmergDate, data=GH_data)

opar <- par(mfrow = c(2, 2))

plot(gh_water_lm)
check_model(gh_water_lm)
water_effects = allEffects(gh_water_lm)

plot(water_effects)
Anova(gh_water_lm)

emmeans(gh_water_lm, list(pairwise~Plant), adjust="tukey")



```

## Linear fat model diagnostic plots

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
