---
title: "NectarExperiment_SummaryRMarkdown"
output:
  word_document: default
  html_document: default
date: "2023-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown Nectar Experiment Analysis Script
### Writen by Manon Veselovsky

```{r, include=FALSE}
rm(list=ls())
```


```{r, include=FALSE}
summarydat<-read.csv("processed/summarydat.csv")
data = summarydat
library(lmerTest)
library(car)
library(effects)
require(ggplot2)
library(emmeans)
library(multcomp)

```

#### create separate working databases for greenhouse and field experiments
```{r}
GH_data = subset(summarydat, summarydat$ExpLoc == "GH")
Field_data = subset(summarydat,summarydat$ExpLoc == "Field")
```

## Data Analysis
### Greenhouse

#### Check the weights on day 7 of trial for normality
```{r, echo=FALSE, fig.cap="Plots to check normality and formal test"}
mygraph <- ggplot(GH_data, aes(x = RawWeight_day7))
mygraph <- mygraph +
  # add data density smooth
  geom_density() +
  # add rug (bars at the bottom of the plot)
  geom_rug() +
  # add black semitransparent histogram
  geom_histogram(aes(y = ..density..),
  color = "black",
  alpha = 0.3) +
  # add normal curve in red, with mean and sd from fklength
  stat_function(fun = dnorm,
  args = list(
  mean = mean(GH_data$RawWeight_day7),
  sd = sd(GH_data$RawWeight_day7)
  ),
  color = "red")

#display graph
mygraph
qqnorm(GH_data$RawWeight_day7)
qqline(GH_data$RawWeight_day7)

# Shapiro formal test for normality
shapiro.test(GH_data$RawWeight_day7)
```

#### Create full model for the raw weight of greenhouse butterflies on day 7 of trial as a function of experimental variables

```{r}
lmer.model = lmer(RawWeight_day7~ Plant + Sex + ForewingLength + EmergDate + Cohort + EnclCol + (1|Couple), data=GH_data)
Anova(lmer.model, type=3)
```

#### Check for collinearity between variables and check model assumptions
```{r, include = FALSE}
vif(lmer.model) #all good
opar <- par(mfrow = c(2, 2))
plot(lmer.model)
# diagnostic plots of the model
qqnorm(resid(lmer.model))
qqline(resid(lmer.model)) #residuals appear normally distributed
```

##### Reduce model, removing least significant terms
```{r, include=FALSE}
# Remove cohort
reduced_1 = lmer(RawWeight_day7~ Plant + Sex + ForewingLength + EmergDate + EnclCol + (1|Couple), data=GH_data)
```

```{r}
#test to see if more complex model is significantly better
anova(reduced_1, lmer.model) #more complex not significantly better, leave cohort out
Anova(reduced_1, type=3) #next least significant term is EmergDate
```

```{r, include=FALSE}
reduced_2 = lmer(RawWeight_day7~ Plant + Sex + ForewingLength + EnclCol + (1|Couple), data=GH_data)
```

```{r}
#test to see if more complex is significantly better
anova(reduced_2, reduced_1) #more complex not significantly better, leave EmergDate out
Anova(reduced_2, type=3) #next least significant is EnclCol
```

```{r, include=FALSE}
reduced_3= lmer(RawWeight_day7~ Plant + Sex + ForewingLength + (1|Couple), data=GH_data)
```

```{r}
#test to see if reduced is significantly better
anova(reduced_3, reduced_2) #more complex is significantly better, keep reduced_2

bestmodel = reduced_2 #assign best model and display its effects plot
```

```{r,echo = FALSE, fig.cap="Effects plot of the best model"}
plot(allEffects(bestmodel))
```

#### Multiple comparisons test to detect differences between plant treatment groups

```{r}
m1<-bestmodel
library(multcomp)
#WHICH groups are different from each other
g<-glht(m1, mcp(Plant="Tukey")); confint(g)
#OR
# library(emmeans)
# emmeans(m1, list(pairwise~Plant), adjust="tukey")
```

### Comparing greenhouse and field goldenrod

```{r, include=FALSE}
#Create a database of field and gh goldenrod to compare experiments
GH_F_data = summarydat[which(summarydat$Plant=="SolAlt"),]

#Check for normality on raw day 7 weights
hist(GH_F_data$RawWeight_day7)

#Use relative weight change (a proportion) in GLMER (binomial distribution for proportion)
gh_f_model = lmer(RawWeight_day7~ Sex + ForewingLength + ExpLoc + (1|Couple), data=GH_F_data)
Anova(gh_f_model, type=3)


plot(allEffects(gh_f_model))
```
### Check flower surface area and if it significantly differs between species
```{r}
SA_comparison = lm(TotalSA ~ Plant + ExpLoc + Plant:ExpLoc, data=data)
emmeans(SA_comparison, list(pairwise~Plant), adjust="tukey")
plot(allEffects(SA_comparison))
```

###
